Variational Monte Carlo on bosonic systems

### 2018-04-03 project_1/article/article.tex
Insignificant changes
Theory revision.

    First, we start by explaining our theoretical quantum system of a bosonic gas trapped
    in a potential well. Here we also outline the principle under which particles in the gas are
    allowed to interact. Moreover, we introduce the concept of \emph{local energy} and
    then quantum \emph{drift force}. Second, we outline in more detailed algebra the
    nature of a non-interaction system. For this system we are able to derive an exact 
    variational energy to which we can compare the results from our simulations. Third,
    we give a description of the particle-particle interaction implemented as if each 
    particle is a hard sphere. Fourth, we provide a thorough description of algorithms and
    statistical methods employed in this study. Here we introduce the Metropolis-Hastings
    algorithm. This algorithm can by a deft trick be expanded with the method of 
    \emph{importance sampling} which make the way in which we pick the next step in
   	the Monte-Carlo cycling much more intelligent. The main point of note with regards to 
   	the statistical analysis is that the data generated by the Monte Carlo cycling is
   	autocorrelated and computation of variance must be done with the \emph{blocking method}.
   	As an automated way of finding the optimal variational parameter, we sketch out the
   	method of \emph{gradient descent}. Fifth, we put the machinere described in the 
   	previous methods to work. We test against the actual values in a non-interacting system
   	before analysing the more interesting perturbed-trap system with interacting particles. 
   	Lastly, we highlight the more interesting aspects of this study in a discussion of the results
   	before ending on some summary remarks.

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        V_{\text{ext}}(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    trap. The two-body Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i = 1}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one body operator, $h$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}_i^2
        + V_{\text{ext}}(\vf{r}_i),
    \end{align}
    (we assume equal mass)
    and the two-body interaction operator, $w$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
    \end{align}
    where $a$ is the hard sphere of the particle. The trial wavefunction,
    $\Psi_T(\vf{r})$, we will be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_{i = 1}^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ and $\beta$ are variational parameters and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\Phi_T(\vf{r})$ the \textit{Slater permanent} consisting of the $N$
    first single particle wavefunctions, and the correlation wavefunction, $f$,
    given by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
    \end{align}
    We will for brevity use the notation $\phi(\vf{r}_i) = \phi_i$ and $r_{jk} =
    |\vf{r}_j - \vf{r}_k|$.

@@

    \subsection{The drift force}
        A disadvantage in the use of the brute-force Metropolis algorithm is
        that we might be spending much computational resources in an
        uninteresting part of configuration space. To make smarter moves we will
        use the Metropolis-Hastings algorithm (which will be discussed in due
        time). This algorithm is dependent on the drift force of the system.
        \begin{align}
            \vf{F}(\vf{r})
            &=
            \sum_{k = 1}^N
            \vf{F}_k(\vf{r})
            =
            \sum_{k = 1}^N
            \frac{2\nabla_k\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
            \label{eq:drift_force}
        \end{align}
        Using this expression we are able to move towards parts of
        configuration space where the gradient increases or decreases yielding a
        better choice of movements. We will mainly be interested in the drift
        force of a single particle $k$.

\section{Non-interacting harmonic oscillators}
    We start by looking at a simple system of non-interacting harmonic
    oscillators. That is, where $a = 0$ and $\beta = 1$. We thus get the trial
    wavefunction
    \begin{align}
        \Psi_T(\vf{r})
        = \Phi_T(\vf{r})
        = \prod_{i = 1}^N \exp\bigl[
            -\alpha |\vf{r}_i|^2
        \bigr],
    \end{align}
    where $|\vf{r}_i| = r_i$. As $a = 0$ the interaction term,
    $w(\vf{r}_i, \vf{r}_j)$, vanishes and the Hamiltonian is given by
    (in the spherical case)
    \begin{align}
        H &= \sum_{i = 1}^N h(\vf{r}_i)
        = \sum_{i = 1}^N \Biggl(
            -\frac{\hbar^2}{2m}\nabla_i^2
            + \half m \omega^2 r_i^2
        \Biggr).
    \end{align}
    To find the drift force and the local energy we have to compute the gradient
    and the Laplacian of the trial wavefunction. The gradient is given by
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
    \end{align}
    whereas the Laplacian yields
    \begin{align}
        \nabla^2_k\Psi_T(\vf{r})
        &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}),
    \end{align}
    where $d$ is the dimensionality of the problem determined by
    $\vf{r}_k \in \mathbb{R}^d$. We can thus use the gradient to find an
    expression for the drift force for particle $k$.
    \begin{align}
        \vf{F}_k(\vf{r})
        &= -4\alpha\vf{r}_k.
    \end{align}
    Using the Laplcian we can compute the kinteic term in the expression for the
    local energy. We get
    \begin{align}
        E_L(\vf{r})
        &=
        \sum_{i = 1}^N
        \Biggl(
            -\frac{\hbar^2}{2m}
            \bigl[
                -2d\alpha + 4\alpha^2 r_i^2
            \bigr]
            + \half m\omega^2 r_i^2
        \Biggr).
    \end{align}
    In natural units, with $\hbar = c = m = 1$, this reduces to
    \begin{align}
        E_L(\vf{r})
        &=
        \alpha dN
        + \biggl(
            \half\omega^2
            - 2\alpha^2
        \biggr)
        \sum_{i = 1}^N r_i^2.
        \label{eq:closed_form_natural_units_local_energy}
    \end{align}
    It is worth noting that for $\alpha = \half\omega$ ($\alpha$ is required to
    be positive) we will find a stable value which turns out to be the exact
    energy minimum. This happens as the entire sum over all the random walkers
    disappears.

    \subsection{Exact variational energy}
        As the system is non-interacting and consisting of Gaussians we can
        find an expression for the exact energy as a function of the
        variational parameter $\alpha$, i.e.,
        \begin{align}
            E(\alpha)
            &=
            \frac{\bra{\Psi_T}H\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}}.
        \end{align}

@@

\section{Interacting hard sphere bosons}
    Moving to the full system allowing $\beta$ to vary and setting $a
    \neq 0$ we can write the trial wavefunction as
    \begin{align}
        \Psi_T(\vf{r})
        &=
        \Phi_T(\vf{r})
        J(\vf{r}),
    \end{align}
    where $\ket{\Phi_T}$ is the same Slater permanent as in
    \autoref{eq:initial_trial_wavefunction} and $J(\vf{r})$ is the
    \textit{Jastrow factor} given by
    \begin{align}
        J(\vf{r})
        &=
        \exp\Biggl(
            \sum_{j < l}^N u(r_{jl})
        \Biggr),
    \end{align}
    where $r_{jk} = |\vf{r}_j - \vf{r}_k|$ and
    \begin{align}
        u(r_{jk}) = \ln\bigl[f(a, \vf{r}_j, \vf{r}_k)\bigr].
    \end{align}
    To further shorten the notation we will use $u_{jk} = u(r_{jk})$. Computing
    the gradient of the wavefunction we get
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &=
        \Bigl[
            \nabla_k
            \Phi_T(\vf{r})
        \Bigr]
        J(\vf{r})
        + \Phi_T(\vf{r})
        \nabla_k J(\vf{r}).
    \end{align}
    The gradient of the Slater permament for particle $k$ is given by
    \begin{align}
        \nabla_k
        \Phi_T(\vf{r})
        &=
        \nabla_k\phi_k
        \prod_{i \neq k}^N\phi_i
        = \frac{\nabla_k\phi_k}{\phi_k}
        \Phi_T(\vf{r}).
    \end{align}
    The gradient of the Jastrow factor is given by
    \begin{align}
        \nabla_k J(\vf{r})
        &=
        J(\vf{r})
        \nabla_k\sum_{m < n}^N u_{mn} \\
        &= J(\vf{r})
        \Biggl(
            \sum_{m = 1}^{k - 1}\nabla_k u_{mk}
            \sum_{n = k + 1}^N\nabla_k u_{kn}
        \Biggr)
        \\
        &=
        J(\vf{r})
        \sum_{m \neq k}^N\nabla_k u_{km},
    \end{align}
    where the gradient of the interaction term splits the antisymmetric
    sum into two parts. As $r_{ij} = r_{ji}$ we can combine these sums
    into a single sum. This in total yields the gradient
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &=
        \Biggl(
            \frac{\nabla_k\phi_k}{\phi_k}
            + \sum_{m \neq k}^N
            \nabla_k u_{km}
        \Biggr)
        \Psi_T(\vf{r}).
        \label{eq:gradient_full_wavefunction}
    \end{align}

    The Laplcian of the trial wavefunction is found by finding the divergence of
    \autoref{eq:gradient_full_wavefunction}.
    \begin{align}
        \nabla_k^2\Psi_T(\vf{r})
        &=
        \Biggl(
            \nabla_k\Biggl[
                \frac{\nabla_k\phi_k}{\phi_k}
            \Biggr]
            +
            \sum_{m \neq k}^N \nabla_k^2 u_{km}
        \Biggr)\Psi_T(\vf{r})
        \\
        &\qquad
        +
        \Biggl(
            \frac{\nabla_k\phi_k}{\phi_k}
            + \sum_{m \neq k}^N
            \nabla_k u_{km}
        \Biggr)^2
        \Psi_T(\vf{r}),
    \end{align}
    where the squared term came from taking the gradient of the trial
    wavefunction.  To further simplify we divide by the trial
    wavefunction. This yields
    \begin{align}
        \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
        &=
        \frac{\nabla_k^2\phi_k}{\phi_k}
        + 2\frac{\nabla_k\phi_k}{\phi_k}
        \sum_{m \neq k}\nabla_k u_{km}
        \nonumber \\
        &\qquad
        + \sum_{m\neq k}^N\nabla_k^2 u_{km}
        + \Biggl(
            \sum_{m \neq k}^N\nabla_k u_{km}
        \Biggr)^2.
    \end{align}
    To go from here we have to find the gradient and the Laplacian of
    the single particle functions, $\phi_k$, and the interaction
    functions $u_{km}$. For the single particle functions we use
    Cartesian coordinates when finding the derivatives whereas we for
    the interaction functions will use spherical coordinates and do a
    variable substitution. Beginning with the gradient of the single
    particle functions we get
    \begin{align}
        \nabla_k\phi_k
        &=
        \nabla_k\exp\bigl[
            -\alpha(x_k^2 + y_k^2 + \beta z_k^2)
        \bigr] \\
        &=
        -2\alpha
        (x_k\vf{e}_i + y_k\vf{e}_j + \beta z_k\vf{e}_k)
        \phi_k.
    \end{align}
    Note that the subscripts on the unit vectors $\vf{e}_i$ are
    \textit{not} the same as the subscripts used for its components. The
    Laplacian yields
    \begin{align}
        \nabla_k^2\phi_k
        &=
        \Bigl[
            -2\alpha
            \big(
                d - 1 + \beta
            \bigr)
            \nonumber \\
            &\qquad
            + 4\alpha^2
            \bigl(
                x_k^2 + y_k^2 + \beta^2z_k^2
            \bigr)
        \Bigr]
        \phi_k,
    \end{align}
    with $d$ as the dimensionality of the problem.
    In order to derive the interaction functions we have to do a
    variable substitution using $r_{km} = |\vf{r}_k - \vf{r}_m|$. We can
    then rewrite the $\nabla_k$-operator as
    \begin{align}
        \nabla_k
        &=
        \nabla_k
        \dpd[]{r_{km}}{r_{km}}
        =
        \nabla_k r_{km} \dpd[]{}{r_{km}}
        \\
        &=
        \frac{\vf{r}_k - \vf{r}_m}{r_{km}}\dpd[]{}{r_{km}}.
    \end{align}
    Applying this version of the $\nabla_k$-operator to $u_{km}$ yields
    \begin{align}
        \nabla_k u_{km}
        &=
        \frac{\vf{r}_k - \vf{r}_m}{r_{km}}
        \dpd[]{u_{km}}{r_{km}}.
    \end{align}

    \begin{widetext}
        For the Laplacian we switch a little back and forth between the
        two ways of representing the $\nabla_k$-operator. We thus get
        \begin{align}
            \nabla_k^2 u_{km}
            &=
            \frac{\nabla_k \vf{r}_k}{r_{km}}\dpd[]{u_{km}}{r_{km}}
            + \Biggl[
                \nabla_k \frac{1}{r_{km}}
            \Biggr]
            (\vf{r}_k - \vf{r}_m)\dpd[]{u_{km}}{r_{km}}
            + \frac{\vf{r}_k - \vf{r}_m}{r_{km}}
            \nabla_k \dpd[]{u_{km}}{r_{km}} \\
            &= \frac{d}{r_{km}}\dpd[]{u_{km}}{r_{km}}
            - \frac{(\vf{r}_k - \vf{r}_m)^2}{r_{km}^3}
            \dpd[]{u_{km}}{r_{km}}
            + \frac{(\vf{r}_k - \vf{r}_m)^2}{r_{km}^2}
            \dpd[2]{u_{km}}{r_{km}} \\
            &=
            \frac{d - 1}{r_{km}}\dpd[]{u_{km}}{r_{km}}
            + \dpd[2]{u_{km}}{r_{km}},
        \end{align}
        where $d$ is again the dimensionality of the problem. In total
        we can state an intermediate version of the Laplacian occuring
        in the local energy as
        \begin{align}
            \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
            &=
            \frac{\nabla_k^2\phi_k}{\phi_k}
            + 2\frac{\nabla_k\phi_k}{\phi_k}
            \sum_{m \neq k}^N
            \frac{\vf{r}_k - \vf{r}_m}{r_{km}}
            \dpd[]{u_{km}}{r_{km}}
            + \sum_{m\neq k}^N
            \Biggl(
                \frac{d - 1}{r_{km}}\dpd[]{u_{km}}{r_{km}}
                + \dpd[2]{u_{km}}{r_{km}}
            \Biggr)
            \nonumber \\
            &\qquad
            +
            \sum_{m, n \neq k}^N
            \frac{\vf{r}_k - \vf{r}_m}{r_{km}}
            \frac{\vf{r}_k - \vf{r}_n}{r_{kn}}
            \dpd[]{u_{km}}{r_{km}}
            \dpd[]{u_{kn}}{r_{kn}}.
        \end{align}
    \end{widetext}
    Moving on to the derivatives of the interaction terms, $u_{km}$, to
    get an explicit expression for the Laplacian.
    \begin{align}
        \dpd[]{u_{km}}{r_{km}}
        &=
        \frac{a}{r_{km}(r_{km} - a)},
        \\
        \dpd[2]{u_{km}}{r_{km}}
        &= \frac{a^2 - 2ar_{km}}{r_{km}^2(r_{km} - a)^2}.
    \end{align}
    The local energy and the drift force can now be found by combining these
    expressions. For brevity, we will not write out the explicit expressions as
    these will be called by separated functions in our programs.

>>>>

    First, we start by explaining our theoretical quantum system of a bosonic gas trapped
    in a potential well. Here we also outline the principle under which particles in the gas are
    allowed to interact. Moreover, we introduce the concept of \emph{local energy} and
    then quantum \emph{drift force}. Here we also outline in more detailed algebra the
    nature of a non-interaction system. For this system we are able to derive an exact 
    variational energy to which we can compare the results from our simulations. Naturally,
    we furthermore give a description of the particle-particle interaction implemented as if each 
    particle is a hard sphere. Second, we provide a thorough description of algorithms and
    statistical methods employed in this study. Here we introduce the Metropolis-Hastings
    algorithm. This algorithm can by a deft trick be expanded with the method of 
    \emph{importance sampling} which make the way in which we pick the next step in
   	the Monte-Carlo cycling much more intelligent. The main point of note with regards to 
   	the statistical analysis is that the data generated by the Monte Carlo cycling is
   	autocorrelated and computation of variance must be done with the \emph{blocking method}.
   	As an automated way of finding the optimal variational parameter, we sketch out the
   	method of \emph{gradient descent}. Third, we put the machinery described in the 
   	previous sections to work. We test against the actual values in a non-interacting system
   	before analysing the more interesting perturbed-trap system with interacting particles. 
   	Lastly, we highlight the more interesting aspects of this study in a discussion of the results
   	before ending on some summary remarks.

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        V_{\text{ext}}(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    trap, where the oscillatory frequency is different in $z$-direction so that
    $\omega_z \neq \omega$. 
    The two-body Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i = 1}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one body operator, $h$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}_i^2
        + V_{\text{ext}}(\vf{r}_i),
    \end{align}
    (we assume equal mass)
    and the two-body interaction operator, $w$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
    \end{align}
    where $a$ is the hard sphere of the particle. The trial wavefunction,
    $\Psi_T(\vf{r})$, we will be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_{i = 1}^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ and $\beta$ are variational parameters and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\Phi_T(\vf{r})$ the \textit{Slater permanent} consisting of the $N$
    first single particle wavefunctions, and the correlation wavefunction, $f$,
    given by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
    \end{align}
    We will for brevity use the notation $\phi(\vf{r}_i) = \phi_i$ and $r_{jk} =
    |\vf{r}_j - \vf{r}_k|$.

@@

    \subsection{The drift force}
        A disadvantage in the use of the brute-force Metropolis-Hastings algorithm is
        that we might be spending much computational resources in an
        uninteresting part of configuration space. To make smarter moves we will
        use the Metropolis-Hastings algorithm with the additions of 
        \emph{Importance Sampling}  (which will be discussed in due
        time). This alteration of the algorithm is dependent on the 
        \emph{drift force} of the system.
        \begin{align}
            \vf{F}(\vf{r})
            &=
            \sum_{k = 1}^N
            \vf{F}_k(\vf{r})
            =
            \sum_{k = 1}^N
            \frac{2\nabla_k\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
            \label{eq:drift_force}
        \end{align}
        Using this expression we are able to move towards parts of
        configuration space where the gradient increases or decreases yielding a
        better choice of movements. We will mainly be interested in the drift
        force of a single particle $k$.

\subsection{Non-interacting harmonic oscillators}
    We start by looking at a simple system of non-interacting harmonic
    oscillators. That is, where $a = 0$ and $\beta = 1$, i.e. there is no hard-shell
    interaction between particles and the particles are confined in a spherical trap.
    We thus get the trial  wavefunction
    \begin{align}
        \Psi_T(\vf{r})
        = \Phi_T(\vf{r})
        = \prod_{i = 1}^N \exp\bigl[
            -\alpha |\vf{r}_i|^2
        \bigr],
    \end{align}
    where $|\vf{r}_i| = r_i$. As $a = 0$ the interaction term,
    $w(\vf{r}_i, \vf{r}_j)$, vanishes and the Hamiltonian is given by
    (in the spherical case)
    \begin{align}
        H &= \sum_{i = 1}^N h(\vf{r}_i)
        = \sum_{i = 1}^N \Biggl(
            -\frac{\hbar^2}{2m}\nabla_i^2
            + \half m \omega^2 r_i^2
        \Biggr).
    \end{align}
    To find the drift force and the local energy we have to compute the gradient
    and the Laplacian of the trial wavefunction. The gradient is given by
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
    \end{align}
    whereas the Laplacian yields
    \begin{align}
        \nabla^2_k\Psi_T(\vf{r})
        &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}),
    \end{align}
    where $d$ is the dimensionality of the problem determined by
    $\vf{r}_k \in \mathbb{R}^d$. We can thus use the gradient to find an
    expression for the drift force for particle $k$.
    \begin{align}
        \vf{F}_k(\vf{r})
        &= -4\alpha\vf{r}_k.
    \end{align}
    Using the Laplacian we can compute the kinetic term in the expression for the
    local energy. We get
    \begin{align}
        E_L(\vf{r})
        &=
        \sum_{i = 1}^N
        \Biggl(
            -\frac{\hbar^2}{2m}
            \bigl[
                -2d\alpha + 4\alpha^2 r_i^2
            \bigr]
            + \half m\omega^2 r_i^2
        \Biggr).
    \end{align}
    In natural units, with $\hbar = c = m = 1$, this reduces to
    \begin{align}
        E_L(\vf{r})
        &=
        \alpha dN
        + \biggl(
            \half\omega^2
            - 2\alpha^2
        \biggr)
        \sum_{i = 1}^N r_i^2.
        \label{eq:closed_form_natural_units_local_energy}
    \end{align}
    It is worth noting that for $\alpha = \half\omega$ ($\alpha$ is required to
    be positive) we will find a stable value which turns out to be the exact
    energy minimum. This happens as the sum over all particles
    disappears.

    \subsubsection{Exact variational energy}
        As the system is non-interacting and consisting of Gaussians we can
        find an expression for the exact energy as a function of the
        variational parameter $\alpha$, i.e.,
        \begin{align}
            E(\alpha)
            &=
            \frac{\bra{\Psi_T}H\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}}.
        \end{align}

@@

\subsection{Interacting hard sphere bosons}
    Moving to the full system allowing $\beta$ to vary and setting $a
    \neq 0$ we can write the trial wavefunction as
    \begin{align}
        \Psi_T(\vf{r})
        &=
        \Phi_T(\vf{r})
        J(\vf{r}),
    \end{align}
    where $\ket{\Phi_T}$ is the same Slater permanent as in
    \autoref{eq:initial_trial_wavefunction} and $J(\vf{r})$ is the
    \textit{Jastrow factor} given by
    \begin{align}
        J(\vf{r})
        &=
        \exp\Biggl(
            \sum_{j < l}^N u(r_{jl})
        \Biggr),
    \end{align}
    where $r_{jk} = |\vf{r}_j - \vf{r}_k|$ and
    \begin{align}
        u(r_{jk}) = \ln\bigl[f(a, \vf{r}_j, \vf{r}_k)\bigr].
    \end{align}
    To further shorten the notation we will use $u_{jk} = u(r_{jk})$. Computing
    the gradient of the wavefunction we get
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &=
        \Bigl[
            \nabla_k
            \Phi_T(\vf{r})
        \Bigr]
        J(\vf{r})
        + \Phi_T(\vf{r})
        \nabla_k J(\vf{r}).
    \end{align}
    The gradient of the Slater permament for particle $k$ is given by
    \begin{align}
        \nabla_k
        \Phi_T(\vf{r})
        &=
        \nabla_k\phi_k
        \prod_{i \neq k}^N\phi_i
        = \frac{\nabla_k\phi_k}{\phi_k}
        \Phi_T(\vf{r}).
    \end{align}
    The gradient of the Jastrow factor is given by
    \begin{align}
        \nabla_k J(\vf{r})
        &=
        J(\vf{r})
        \nabla_k\sum_{m < n}^N u_{mn} \\
        &= J(\vf{r})
        \Biggl(
            \sum_{m = 1}^{k - 1}\nabla_k u_{mk}
            \sum_{n = k + 1}^N\nabla_k u_{kn}
        \Biggr)
        \\
        &=
        J(\vf{r})
        \sum_{m \neq k}^N\nabla_k u_{km},
    \end{align}
    where the gradient of the interaction term splits the antisymmetric
    sum into two parts. As $r_{ij} = r_{ji}$ we can combine these sums
    into a single sum. This in total yields the gradient
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &=
        \Biggl(
            \frac{\nabla_k\phi_k}{\phi_k}
            + \sum_{m \neq k}^N
            \nabla_k u_{km}
        \Biggr)
        \Psi_T(\vf{r}).
        \label{eq:gradient_full_wavefunction}
    \end{align}

    The Laplcian of the trial wavefunction is found by finding the divergence of
    \autoref{eq:gradient_full_wavefunction}.
    \begin{align}
        \nabla_k^2\Psi_T(\vf{r})
        &=
        \Biggl(
            \nabla_k\Biggl[
                \frac{\nabla_k\phi_k}{\phi_k}
            \Biggr]
            +
            \sum_{m \neq k}^N \nabla_k^2 u_{km}
        \Biggr)\Psi_T(\vf{r})
        \\
        &\qquad
        +
        \Biggl(
            \frac{\nabla_k\phi_k}{\phi_k}
            + \sum_{m \neq k}^N
            \nabla_k u_{km}
        \Biggr)^2
        \Psi_T(\vf{r}),
    \end{align}
    where the squared term came from taking the gradient of the trial
    wavefunction.  To further simplify we divide by the trial
    wavefunction. This yields
    \begin{align}
        \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
        &=
        \frac{\nabla_k^2\phi_k}{\phi_k}
        + 2\frac{\nabla_k\phi_k}{\phi_k}
        \sum_{m \neq k}\nabla_k u_{km}
        \nonumber \\
        &\qquad
        + \sum_{m\neq k}^N\nabla_k^2 u_{km}
        + \Biggl(
            \sum_{m \neq k}^N\nabla_k u_{km}
        \Biggr)^2.
    \end{align}
    From here, the next step is to find the gradient and the Laplacian of
    the single particle functions, $\phi_k$, and the interaction
    functions $u_{km}$. For the single particle functions we use
    Cartesian coordinates when finding the derivatives while for
    the interaction functions we use spherical coordinates and do a
    variable substitution. Beginning with the gradient of the single
    particle functions we get
    \begin{align}
        \nabla_k\phi_k
        &=
        \nabla_k\exp\bigl[
            -\alpha(x_k^2 + y_k^2 + \beta z_k^2)
        \bigr] \\
        &=
        -2\alpha
        (x_k\vf{e}_i + y_k\vf{e}_j + \beta z_k\vf{e}_k)
        \phi_k.
    \end{align}
    Note that the subscripts on the unit vectors $\vf{e}_i$ are
    \textit{not} the same as the subscripts used for its components. The
    Laplacian yields
    \begin{align}
        \nabla_k^2\phi_k
        &=
        \Bigl[
            -2\alpha
            \big(
                d - 1 + \beta
            \bigr)
            \nonumber \\
            &\qquad
            + 4\alpha^2
            \bigl(
                x_k^2 + y_k^2 + \beta^2z_k^2
            \bigr)
        \Bigr]
        \phi_k,
    \end{align}
    with $d$ as the dimensionality of the problem.
    In order to derive the interaction functions we have to do a
    variable substitution using $r_{km} = |\vf{r}_k - \vf{r}_m|$. We can
    then rewrite the $\nabla_k$-operator as
    \begin{align}
        \nabla_k
        &=
        \nabla_k
        \dpd[]{r_{km}}{r_{km}}
        =
        \nabla_k r_{km} \dpd[]{}{r_{km}}
        \\
        &=
        \frac{\vf{r}_k - \vf{r}_m}{r_{km}}\dpd[]{}{r_{km}}.
    \end{align}
    Applying this version of the $\nabla_k$-operator to $u_{km}$ yields
    \begin{align}
        \nabla_k u_{km}
        &=
        \frac{\vf{r}_k - \vf{r}_m}{r_{km}}
        \dpd[]{u_{km}}{r_{km}}.
    \end{align}

    \begin{widetext}
        For the Laplacian we switch a little back and forth between the
        two ways of representing the $\nabla_k$-operator. We thus get
        \begin{align}
            \nabla_k^2 u_{km}
            &=
            \frac{\nabla_k \vf{r}_k}{r_{km}}\dpd[]{u_{km}}{r_{km}}
            + \Biggl[
                \nabla_k \frac{1}{r_{km}}
            \Biggr]
            (\vf{r}_k - \vf{r}_m)\dpd[]{u_{km}}{r_{km}}
            + \frac{\vf{r}_k - \vf{r}_m}{r_{km}}
            \nabla_k \dpd[]{u_{km}}{r_{km}} \\
            &= \frac{d}{r_{km}}\dpd[]{u_{km}}{r_{km}}
            - \frac{(\vf{r}_k - \vf{r}_m)^2}{r_{km}^3}
            \dpd[]{u_{km}}{r_{km}}
            + \frac{(\vf{r}_k - \vf{r}_m)^2}{r_{km}^2}
            \dpd[2]{u_{km}}{r_{km}} \\
            &=
            \frac{d - 1}{r_{km}}\dpd[]{u_{km}}{r_{km}}
            + \dpd[2]{u_{km}}{r_{km}},
        \end{align}
        where $d$ is again the dimensionality of the problem. In total
        we can state an intermediate version of the Laplacian occuring
        in the local energy as
        \begin{align}
            \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
            &=
            \frac{\nabla_k^2\phi_k}{\phi_k}
            + 2\frac{\nabla_k\phi_k}{\phi_k}
            \sum_{m \neq k}^N
            \frac{\vf{r}_k - \vf{r}_m}{r_{km}}
            \dpd[]{u_{km}}{r_{km}}
            + \sum_{m\neq k}^N
            \Biggl(
                \frac{d - 1}{r_{km}}\dpd[]{u_{km}}{r_{km}}
                + \dpd[2]{u_{km}}{r_{km}}
            \Biggr)
            \nonumber \\
            &\qquad
            +
            \sum_{m, n \neq k}^N
            \frac{\vf{r}_k - \vf{r}_m}{r_{km}}
            \frac{\vf{r}_k - \vf{r}_n}{r_{kn}}
            \dpd[]{u_{km}}{r_{km}}
            \dpd[]{u_{kn}}{r_{kn}}.
        \end{align}
    \end{widetext}
    Moving on to the derivatives of the interaction terms, $u_{km}$, to
    get an explicit expression for the Laplacian.
    \begin{align}
        \dpd[]{u_{km}}{r_{km}}
        &=
        \frac{a}{r_{km}(r_{km} - a)},
        \\
        \dpd[2]{u_{km}}{r_{km}}
        &= \frac{a^2 - 2ar_{km}}{r_{km}^2(r_{km} - a)^2}.
    \end{align}
    The local energy and the drift force can now be found by combining these
    expressions. We will not write out the explicit expressions as
    these will be called by separated functions in our programs.

    \subsubsection{Scaling the system}
        We now introduce a scaled distance $\vf{r}' = \vf{r}/a_{\text{ho}}$
        \cite{dubois2001bose}, where
        \begin{align}
            a_{\text{ho}}
            &=
            \sqrt{\frac{\hbar}{m\omega}}.
        \end{align}
        We can then rewrite the Hamiltonian for the elliptic potential in terms
        of this new distance. By doing a variable substitution for each
        direction in the Laplace operator we get
        \begin{align}
            \nabla_k^2
            &=
            \frac{1}{a_{\text{ho}}^2}{\nabla_k'}^2.
        \end{align}
